---
title: "EDA practical"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We are going to use the iris data in this practical.

## Plotting data

Use `ggplot` to make a plot showing the distribution of petal widths for the three species.

There are several valid ways to do this. Some possibly useful geoms include

```{r, eval = FALSE}
geom_histogram
geom_density
geom_boxplot
geom_violin
geom_jitter
ggforce::geom_sina # you will need to install the package
```

Remember, you can use several geoms together.

Choose a plot that you think best represents the data.


## Univariate summary statistics

Calculate the mean and standard deviation for petal width for each species. `dplyr::summarise` might be useful.

## Bivariate analyses

Use `GGally::ggpairs` to plot all variables in the iris data against each other.

Find the covariance between sepal length and petal length for each species.

Calculate the correlation between sepal length and petal length from the covariance and variance. Check this matches the correlation calculated directly with `cor`.

Are the correlations significant for each species? Again, you can do this with `summarise`, with `cor.test` help from `broom::tidy`.
