---
title: "Exploratory Data Analysis"
author: Richard J. Telford
output: ioslides_presentation
---

<style>
.forceBreak { -webkit-column-break-after: always; break-after: column; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("patchwork")
theme_set(theme_bw())
```

## Exploratory Data Analysis

Distinct from confirmatory data analysis

Range of predominantly graphical tools

- Hypotheses generation
- Outlier detection
- Suggest appropriate data transformations

Essential first step of any analysis

## Exploratory + Confirmatory Analysis

- Risk of data dredging

  - Searching for interesting patterns then testing if they are statistically significant   
  - Raised risk of falsely finding statistically significant result  
  
- Well designed study has two phases  
  - Gather small initial data set   
  - Use exploratory data analysis on initial dataset to determine features of interest and statistical protocol  
  - Gather new data and test hypotheses with protocol developed in first phase  
  - Can also split large dataset into two parts 
  
# Univariate EDA  
  
  
## Summarising data

Distribution - the shape of the data

```{r, echo = FALSE, message = FALSE}
hist <- ggplot(iris, aes(x = Sepal.Length)) +
  geom_histogram(bins = 20)
den <- ggplot(iris, aes(x = Sepal.Length)) +
  geom_density()

hist + den
```


## Central point

- Mean
  - $\mu = n^{-1}\Sigma y_i = \Sigma p(y_i)y_i$

```{r}
mean(iris$Sepal.Length)
```

- Median
  - Sort from min to max
  - The observation in the middle is the median

Median is robust to outliers
  
```{r}
median(iris$Sepal.Length)
```
  
## Quantiles/percentiles

Median = second quantile = 50^th^ percentilef

Other quantiles/percentiles are available

```{r}
quantile(iris$Sepal.Length, probs = c(0.25, 0.50, 0.75, 0.975))
```

```{r, fig.height = 3, fig.width = 4, echo = FALSE}
ggplot(iris, aes(x = Sepal.Length)) +
  geom_histogram(bins = 20) +
  geom_vline(xintercept = quantile(iris$Sepal.Length, probs = c(0.25, 0.50, 0.75, 0.975)))
```

## Spread

Spread of observations around a central point.

- Absolute difference $s = \Sigma|y_i - \mu|$
- Variance $s^2 = (n-1)^{-1}\Sigma (y_i - \mu)^2$
  - mean squared difference around the mean
- Standard deviation $s = \sqrt {s^2}$

## Standard deviation $s = \sqrt{(n-1)^{-1}\Sigma (y_i - \mu)^2}$

- $\mu \pm 1s$ = 68% of observations
- $\mu \pm 1.96s$ = 95% of observations
- $\mu \pm 1s$ point of inflection

```{r, fig.height = 3, fig.width = 4, echo = FALSE}
ggplot(iris, aes(x = Sepal.Length)) +
  geom_histogram(bins = 20, aes(y = stat(density))) +
  stat_function(fun = dnorm, args = list(mean = mean(iris$Sepal.Length), sd = sd(iris$Sepal.Length)), colour = "blue") +
  geom_vline(xintercept = mean(iris$Sepal.Length) + c(1, -1) * sd(iris$Sepal.Length), colour = "red", linetype = "dashed") +
  geom_vline(xintercept = mean(iris$Sepal.Length), colour = "red")
```

## Summary of data

Five number summary
```{r}
summary(iris$Sepal.Length)
```

Standard deviation
```{r}
sqrt(var(iris$Sepal.Length))
sd(iris$Sepal.Length)
```

## Standard deviation vs Standard error {.columns-2 .smaller}

### Standard deviation

$s = \sqrt{(n-1)^{-1}\Sigma (y_i - \mu)^2}$

Spread of observations about mean

$\mu \pm 1s$

```{r, echo = FALSE}
mean(iris$Sepal.Length) + c(-1, 1) * sd(iris$Sepal.Length)
```

Calculate with
```{r, eval = FALSE}
mean(x) + sd(x)
mean(x) - sd(x)
```



### Standard error

$se = s/\sqrt{n}$

Uncertainty of mean

$\mu \pm 1se$

```{r, echo = FALSE}
mean(iris$Sepal.Length) + c(-1, 1) * sd(iris$Sepal.Length) / sqrt(length(iris$Sepal.Length))
```

Need to calculate 

```{r, eval = FALSE}
mean(x) + sd(x) / sqrt(length(x))
mean(x) - sd(x) / sqrt(length(x))
```

## Categorical Variables {.smaller}

Best to represent as a factor (can set sort order)

```{r}
mpg <- mpg %>% mutate(class = factor(class))
```

Number of levels

```{r}
nlevels(mpg$class)
```

Frequency

```{r}
table(mpg$class)
# or count() if column in data frame
```

```{r}
table(mpg$class)
# or count() if column in data frame
```

##

```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar()
```

# Bivariate EDA

## Summaries of several variables {.columns-2}

- Individual summaries
- Combined summaries - measures of association
  - covariance
  - correlation
  
<p class="forceBreak"></p>

```{r, fig.width = 3.5, fig.height = 5, echo = FALSE}
g1 <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

g2 <- g1 + aes(x = Sepal.Width, y = Sepal.Length)

g1/g2

```


## Covariance

Association between two variables

$S_{xy} = \frac{\Sigma (x_i - \mu_x)(y_i - \mu_y)}{n - 1}$

$S_{xy} = S_{yx}$

- \-inf, 0, +inf
- \+ = positive association
- \- = negative association

```{r}
var(iris$Sepal.Length, iris$Sepal.Width)
cov(iris$Sepal.Length, iris$Sepal.Width)
```

## Correlation {.smaller}

Standardised association

- \-1, 0, +1
- \+ = positive association
- \- = negative association

- $S_{xy}$ - covariance 
- $S_x^2$ - variance of x
- $S_y^2$ - variance of y

$r_{xy} = \frac{S_{xy}}{\sqrt{S_x^2S_y^2}}$

$r_{xy} = r_{yx}$


```{r}
cor(iris$Sepal.Length, iris$Sepal.Width)
```

## Covariance vs Correlation {.columns-2 .smaller}

### Covariance

- -inf, 0, +inf
- scale dependent
- not testable

<p class="forceBreak"></p>

### Correlation

- -1, 0, +1
- scale independent
- testable

```{r}
cor.test(iris$Sepal.Length, iris$Sepal.Width)
```

## Covariance and Correlation matrices

```{r}
df <- select(iris, Sepal.Width, Sepal.Length)
var(df)
```

```{r}
cor(df)
```

## Summary of functions used

```{r eval = FALSE}
summary()
mean()
median()
var()
sd()
cor()
cor.test()
```

